\chapter{Items}

This chapter describes syntax and informal semantics of Intentio items. An \emph{item}\index{item} is a component of a module. Items are uniquely named within module and are organized in flat structure.

\begin{bnf}
  \gd{itemdecl} \eq    \gp{fundecl}    \gcomm{function}
                \gorln \gp{importdecl} \gcomm{import declaration}
\end{bnf}

\section{Functions}

\begin{bnf}
  \gd{fundecl} \eq \term{fun} \ \term{\gp{id}} \ \term{(} \ \gtry{\gp{funparams}} \ \term{)} \ \gp{braceblock} \\
  \gd{funparams} \eq \term{\gp{expr}} \ \gmany{ \ \term{,} \ \term{expr} \ \term{\gp{expr}} \ } \ \gtry{\term{,}}
\end{bnf}

A \emph{function}\index{item!function} is a named block, along with optional set of parameters. Functions are declared with the keyword \lstinline{fun}. If keyword \emph{return} is not used, function returns the result of evaluating last expression from contained block. Functions may declare a set of input variables as parameters, through which the caller passes arguments into the function. Input variables behave the same as normal variables defined within function block.

A keyword \lstinline{expr} before argument means that it would be evaluated. If it fails, the function will not be executed and return fail unit term.

The number of function parameters $n$ is called the \emph{arity}\index{items!function arity} of the function.

\begin{example}[Simple function]
\begin{lstlisting}
fun the_hardest_calculation(x) {
  x * 2
}
\end{lstlisting}
\end{example}

\section{Import declarations}

\begin{bnf}
  \gp{importdecl} \eq \term{import} \ \gp{importspec} \\
  \\
  \gp{importspec} \eq    \term{\gp{id}} \gcomm{qualified import}
                  \gorln \term{\gp{id}} \ \term{as} \ \term{\gp{id}} \gcomm{renamed qualified import}
                  \gorln \term{\gp{id}} \ \term{:} \ \term{\gp{id}} \gcomm{item import}
                  \gorln \term{\gp{id}} \ \term{:} \ \term{\gp{id}} \ \term{as} \ \term{\gp{id}} \gcomm{renamed item import}
                  \gorln \term{\gp{id}} \ \term{:} \ \term{*} \gcomm{import-all}
\end{bnf}

An \emph{import declaration}\index{item!import declaration} creates local name binding in current scope, synonymous with specified module. Imports can import exported items directly into current scope or they can be \emph{qualified} where whole module is imported and exported items can be accessed using \emph{qualified identifier} syntax.

Import declarations support five forms:

\begin{description}
  \item [Qualified import] \hfill \\
    Binding target module to its name: \\
    \lstinline{import a}
  \item [Renamed qualified import] \hfill \\
    Binding target module to new local name: \\
    \lstinline{import a as b}
  \item [Item import] \hfill \\
    Binding exported item of target module to its name: \\
    \lstinline{import a:x}
  \item [Renamed item import] \hfill \\
    Binding exported item of target module to new local name: \\
    \lstinline{import a:x as y}
  \item [Import-all] \hfill \\
    Binding all exported items of target module to their names: \\
    \lstinline{import a:*}
\end{description}

An import declaration declares a dependency relation between the importing and imported module. Modules may be mutually recursive, though module must not import itself, directly or indirectly.

\begin{example}[Import declarations]
\begin{lstlisting}
import io
import math:sin
import math as m

fun main() {
  f = io:open("result.txt", "w");
  io:writeln(f, sin(m:pi));
}
\end{lstlisting}
\end{example}
